Preparing your manuscript

### 2024-03-15 paper/paper.qmd
update paper draft

plain-language-summary: |
  A Plain Language Summary (PLS) can be an incredibly effective science communication tool. By summarizing your paper in non-technical terms, you can explain your research and its relevance to a much broader audience. A PLS is required for submissions to AGU Advances, G-Cubed, GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets, JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of Geophysics, but optional for other journals.  A PLS should be no longer than 200 words and should be free of jargon, acronyms, equations, and any technical information that would be unknown to people from outside your scientific discipline. Read our tips for creating an effective PLS.
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: bibliography.bib  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---
## Introduction

River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas [@hagemannHighResolutionDischarge2020]. For the Baltic Sea river runoff is even more important, as the Baltic Sea is nearly decoupled from the open ocean (see Figure). As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers. More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface. The net precipitation accounts for 11 \% and the river input for 89 \% of the total freshwater input (Meier and Doescher, 2002). Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation.

In the context of climate change studies, river runoff is usually generated in two ways. First, river runoff as input for ocean models can be created using hyrdological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020]. HD calculates the water balance using hydrological processes (e.g. snow, glaciers, soil moisture, groundwater contribution). It represents a complex forecast tool that uses underlying physical processes. A different approach would use data-based models that intergrate statistical correction, using the land surface schemes of global or regional climate models.

The releatively recent rise of machine learning (ML) models has been mostly explored for river runoff forcasting, as accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management [@fangExaminingApplicabilityDifferent2019;@tanAdaptiveMiddleLongterm2018;@yangDerivingOperatingRules2018]. Common approach us employ artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks and have gained traction for long-term hydrological forecasting due to their commendable performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).

LSTM networks, an evolution of the classical Recurrent Neural Networks (RNNs), have shown stability and efficacy in sequence-to-sequence predictions, such as using climatic indices for rainfall estimation or long-term hydrological forecasting. However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel.

Recognizing this, we use the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from atmospheric fields for predicting river runoff in the Baltic Sea catchment, summarized by 97 inidivual rivers.

Following the previous description of the complex hyrdological model E-Hype - when present E-Hype should be used. However, in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, RNNs represent a robust way to predict river runoff for any give period of time using atmospheric forcing.

Modeling the Baltic Sea is to a large part the result of the quality of the freshwater input, that is used for the simulation. Meier and Kauker(2003) showed that decadal salinity variations of about 1 $g$ $kg^{-1}$ are caused, inter alia, by annual runoff variations. Further, Meier and Kauker (2003) showed that about 50 \% of the decadal salinity variability can be explained by variations in freshwater input into the Baltic Sea.

This paper delves into the application of deep learning, particularly ConvLSTM, to the challenging task of precipitation nowcasting, a domain yet to fully harness the potential of advanced machine learning techniques. We present ConvLSTM as a novel solution to this spatiotemporal sequence forecasting challenge, highlighting its advantages and potential future applications.

@@

![Inner structure of a Long Short-Term Memory Cell](draw_lstm.pdf){#fig-lstm}

@@

- $x_t$: Input vector at time step $t$.
- $h_{t-1}$: Hidden state from the previous time step.
- $C_{t-1}$: Cell state from the previous time step.
- $W$ and $b$: Weight matrices and bias vectors, respectively, associated with the gates of the LSTM. The subscripts denote the specific gate or operation they are associated with (e.g., $W_f$ and $b_f$ are the weight matrix and bias for the forget gate).
- $\sigma$: Sigmoid activation function ($0,\ldots 1$)
- $\tanh$: Hyperbolic tangent activation function ($ -1 \ldots 1$)

@@

We first implemented the ConvLSTM using and encoder/decoder structure as discussed in ... . To predict all 97 rivers entering the Baltic Sea at once, we flatten the output and use fully connected layers resulting in 97 outputs.

@@

| Parameter name | Parameter size | 
|------|------|
| Channel size | 4 |
| Num. hidden layer |10 |
| Num. timesteps     | 30    |
| Conv. Kernelsize    | (9,9)    |
| Num. ConvLSTM layers    | 1    |
| Batch size| 64 |
| Learning Rate | 1e-3 with CosineAnnealing |

@@

@@

:::{#refs}

:::

>>>>

plain-language-summary: |
  A Plain Language Summary (PLS) can be an incredibly effective science communication tool. By summarizing your paper in non-technical terms, you can explain your research and its relevance to a much broader audience. A PLS is required for submissions to AGU Advances, G-Cubed, GeoHealth, GRL, JAMES, JGR: Biogeosciences, JGR: Oceans, JGR: Planets, JGR: Solid Earth, JGR: Atmospheres, Space Weather, and Reviews of Geophysics, but optional for other journals.  A PLS should be no longer than 200 words and should be free of jargon, acronyms, equations, and any technical information that would be unknown to people from outside your scientific discipline. Read our tips for creating an effective PLS.
keywords: []
key-points:
  - Key Points convey the main points and conclusions of the article. 
  - Up to three key point statements are allowed, and each is limited to at most 140 characters with no abbreviations.
  - Key Points are not included in the word count.
bibliography: bibliography.bib  
citation:
  container-title: Geophysical Research Letters
keep-tex: true
date: last-modified
---

## Introduction

River runoff is an important component of the global water cycle as it comprises about one third of the precipitation over land areas [@hagemannHighResolutionDischarge2020]. In the context of climate change studies, river runoff is usually generated in two ways. First, river runoff as input for ocean models can be created using hyrdological models such as the Hydrological Discharge (HD) model [@hagemannHighResolutionDischarge2020]. HD models calculates the water balance using hydrological processes (e.g. snow, glaciers, soil moisture, groundwater contribution). It represents a complex forecasting tool that uses underlying physical processes. A different approach would use data-based models that intergrate statistical correction, using the land surface schemes of global or regional climate models.

The releatively recent rise of machine learning (ML) models has been mostly explored for river runoff forecasting, as accurate runoff forecasting, especially over extended periods, is pivotal for effective water resources management [@fangExaminingApplicabilityDifferent2019; @tanAdaptiveMiddleLongterm2018; @yangDerivingOperatingRules2018]. Common approaches employ artificial neural networks, support vector machines, adaptive neuro-fuzzy inference systems, and notably, Long Short-Term Memory (LSTM) neural networks that have gained traction for long-term hydrological forecasting due to their excellent performance (Humphrey et al 2016, Huang et al 2014, Ashrafi et al 2017, Yuan et al 2018, Xu et al 2021).

LSTM networks, an evolution of the classical Recurrent Neural Networks (RNNs), have shown stability and efficacy in sequence-to-sequence predictions, such as using climatic indices for rainfall estimation or long-term hydrological forecasting. However, a limitation of LSTMs is their inability to effectively capture two-dimensional structures, an area where Convolutional Neural Networks (CNNs) excel. Recognizing this, we use the ConvLSTM, which integrates the strengths of both LSTM and CNN, to extract spatiotemporal features from atmospheric fields for predicting river runoff.

TODO: More about ConvLSTM

In the following we will use the Baltic Sea catchment to illustrate our approach, while in principal the methodology we propose is universally applicable across various geographic regions. The Baltic Sea serves as a challenging example due to its unique hydrological characteristics, being nearly decoupled from the open ocean (see Figure).

As a consequence, the salinity of the Baltic Sea is driven to a large part by freshwater supply from rivers. More generally, the freshwater input into the Baltic Sea comes either as river runoff or a positive net precipitation (precipitation minus evaporation) over the sea surface. The net precipitation accounts for 11 % and the river input for 89 % of the total freshwater input (Meier and Doescher, 2002). Modeling the Baltic Sea is therefore to a large part the result of the quality of the river input, that is used for the simulation. This makes the accurate modeling of river runoff especially critical for simulations pertaining to the Baltic Sea. In the following we show that in absence of a fully functioning hydrogolocial model, that also uses a rather complex parametrization, ConvLSTM represent a robust way to predict river runoff for any given period of time using only atmospheric forcing.

In this work, we present a ConvLSTM architecture as a novel solution to this spatiotemporal sequence forecasting challenge, highlighting its advantages and potential future applications.

@@

![Inner structure of a Long Short-Term Memory Cell](draw_lstm.png){#fig-lstm fig-align="center"}

@@

-   $x_t$: Input vector at time step $t$.
-   $h_{t-1}$: Hidden state from the previous time step.
-   $C_{t-1}$: Cell state from the previous time step.
-   $W$ and $b$: Weight matrices and bias vectors, respectively, associated with the gates of the LSTM. The subscripts denote the specific gate or operation they are associated with (e.g., $W_f$ and $b_f$ are the weight matrix and bias for the forget gate).
-   $\sigma$: Sigmoid activation function ($0,\ldots 1$)
-   $\tanh$: Hyperbolic tangent activation function (\$ -1 \ldots 1\$)

@@

The ConvLSTM architectures uses and encoder/decoder structure as discussed in TODO. To predict all 97 rivers entering the Baltic Sea, we flatten the output and use fully connected layers to map onto the individual rivers outputs.

@@

| Parameter name       | Parameter size            |
|----------------------|---------------------------|
| Channel size         | 4                         |
| Num. hidden layer    | 10                        |
| Num. timesteps       | 30                        |
| Conv. Kernelsize     | (9,9)                     |
| Num. ConvLSTM layers | 1                         |
| Batch size           | 64                        |
|  Learning Rate       | 1e-3 with CosineAnnealing |

@@

As input the model receives 30 days of atmospheric surface fields temperature $T$, precipitation $P$, specific humidity $Q$ and wind speed $W$, with a daily resolution to predict the river runoff $R$ at the time step $\Delta t+1$ , which can be summarized as

$R_{\Delta t+1} = f\left(T_{t-30:t}, P_{t-30:t}, Q_{t-30:t}, W_{t-30:t}\right)$

with f being a function maps the 30 days of daily atmospheric surface fields data to the predicted river runoff.

The choice of atmospheric fields was based on the implemented river runoff calculation in the atmospheric model COSMO-CLM which uses these four fields to calculate an river runoff estimate.

## Results

For the evaluation of the model performance we consider the period 1979 to 2011. For this period no bias correction was applied to the orignal E-HYPE dataset. We chose a split of 80% training data, 10% validation data to evaluate the performance of the model during training, and 10% training data that is finally used to evaluate the performance of the model after training.

@fig-PerformanceNeuralNetworkRunoff shows the perfomance of the model using the test dataset. The predicted total river runoff for the Baltic Sea is closely matching the original data. Zooming in on the largest individual rivers (lower panels) it can be seen that that also the prediction of the inidivual rivers is close to the original data.

![Figure ..](images/paste-1.png){#fig-PerformanceNeuralNetworkRunoff}

The accuracy of the model is further displayed in @fig-statistical-evaluationNN . The correlation is close to 1 and the residuals show the error of the model is below 1% of the original data.

![Figure ...](images/paste-4.png){#fig-statistical-evaluationNN}

Lastly, we evaluated the performance of the runoff model by incorporating the predicted river runoff as forcing into the ocean model MOM5. This provides a robust validation of the runoff model against more complex real world conditions. This allows us to ensure that the predictions accurately reflect the impact of the river discharge on the ocean dynamics, validating the temporal and spatial variability of the the river discharge. @fig-by15 shows the salinity comparison between the original E-HYPE river runoff and the predicted river runoff at BY15 - a central stations in the Baltic Sea.

![](images/paste-5.png){#fig-by15}

@@

::: {#refs}
:::

### 2024-03-15 paper/paper.tex
update paper draft

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

{\centering \includegraphics{draw_lstm.pdf}

@@

\begin{itemize}
\tightlist
\item
  \(x_t\): Input vector at time step \(t\).
\item
  \(h_{t-1}\): Hidden state from the previous time step.
\item
  \(C_{t-1}\): Cell state from the previous time step.
\item
  \(W\) and \(b\): Weight matrices and bias vectors, respectively,
  associated with the gates of the LSTM. The subscripts denote the
  specific gate or operation they are associated with (e.g., \(W_f\) and
  \(b_f\) are the weight matrix and bias for the forget gate).
\item
  \(\sigma\): Sigmoid activation function (\(0,\ldots 1\))
\item
  \(\tanh\): Hyperbolic tangent activation function (\$-1 \ldots1)
\end{itemize}

@@

We first implemented the ConvLSTM using and encoder/decoder structure as
discussed in \ldots{} . To predict all 97 rivers entering the Baltic Sea
at once, we flatten the output and use fully connected layers resulting
in 97 outputs.

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, interior hidden, boxrule=0pt, sharp corners, enhanced]}{\end{tcolorbox}}\fi

@@

{\centering \includegraphics{draw_lstm.png}

@@

\begin{itemize}
\tightlist
\item
  \(x_t\): Input vector at time step \(t\).
\item
  \(h_{t-1}\): Hidden state from the previous time step.
\item
  \(C_{t-1}\): Cell state from the previous time step.
\item
  \(W\) and \(b\): Weight matrices and bias vectors, respectively,
  associated with the gates of the LSTM. The subscripts denote the
  specific gate or operation they are associated with (e.g., \(W_f\) and
  \(b_f\) are the weight matrix and bias for the forget gate).
\item
  \(\sigma\): Sigmoid activation function (\(0,\ldots 1\))
\item
  \(\tanh\): Hyperbolic tangent activation function (\$ -1 \ldots 1\$)
\end{itemize}

@@

The ConvLSTM architectures uses and encoder/decoder structure as
discussed in TODO. To predict all 97 rivers entering the Baltic Sea, we
flatten the output and use fully connected layers to map onto the
individual rivers outputs.

@@

As input the model receives 30 days of atmospheric surface fields
temperature \(T\), precipitation \(P\), specific humidity \(Q\) and wind
speed \(W\), with a daily resolution to predict the river runoff \(R\)
at the time step \(\Delta t+1\) , which can be summarized as

\(R_{\Delta t+1} = f\left(T_{t-30:t}, P_{t-30:t}, Q_{t-30:t}, W_{t-30:t}\right)\)

with f being a function maps the 30 days of daily atmospheric surface
fields data to the predicted river runoff.

The choice of atmospheric fields was based on the implemented river
runoff calculation in the atmospheric model COSMO-CLM which uses these
four fields to calculate an river runoff estimate.

\hypertarget{results}{%
\section{Results}\label{results}}

For the evaluation of the model performance we consider the period 1979
to 2011. For this period no bias correction was applied to the orignal
E-HYPE dataset. We chose a split of 80\% training data, 10\% validation
data to evaluate the performance of the model during training, and 10\%
training data that is finally used to evaluate the performance of the
model after training.

Figure~\ref{fig-PerformanceNeuralNetworkRunoff} shows the perfomance of
the model using the test dataset. The predicted total river runoff for
the Baltic Sea is closely matching the original data. Zooming in on the
largest individual rivers (lower panels) it can be seen that that also
the prediction of the inidivual rivers is close to the original data.

{\centering \includegraphics{images/paste-1.png}

\caption{\label{fig-PerformanceNeuralNetworkRunoff}Figure ..}

The accuracy of the model is further displayed in
Figure~\ref{fig-statistical-evaluationNN} . The correlation is close to
1 and the residuals show the error of the model is below 1\% of the
original data.

{\centering \includegraphics{images/paste-4.png}

\caption{\label{fig-statistical-evaluationNN}Figure \ldots{}}

Lastly, we evaluated the performance of the runoff model by
incorporating the predicted river runoff as forcing into the ocean model
MOM5. This provides a robust validation of the runoff model against more
complex real world conditions. This allows us to ensure that the
predictions accurately reflect the impact of the river discharge on the
ocean dynamics, validating the temporal and spatial variability of the
the river discharge. \textbf{?@fig-by15} shows the salinity comparison
between the original E-HYPE river runoff and the predicted river runoff
at BY15 - a central stations in the Baltic Sea.

\includegraphics{images/paste-5.png}
